A Formal Model of Checked C

### 2022-07-15 sandp23/formal.tex
update.

\begin{figure}
  \small \centering
  \[ \hspace*{-1em}
\begin{array}{l}
\begin{array}{ll}
       \text{Variables:}~ x
& \text{Integers:}~n::=\mathbb{Z} 
\end{array}
\\[0.5em]

@@

\text{Expression:}& e & ::= & 
\evalue{n}{\tau} \mid x \mid \ecast{\tau}{e} \mid \edyncast{\tau}{e} \mid \ebinop{e}{e} \\[0.2em]
&&\mid& \emalloc{\omega} \mid \estrlen{x} \mid\estar{e}\mid\eassign{e}{e}  \\[0.2em]
&&\mid& \elet{x}{e}{e} \mid \eif{e}{e}{e}
\\[0.2em]
&&\mid&  \ecall{e}{\overline{e}} \mid
\echecked{\overline{x}}{\tau}{e}
\\[0.2em]
&&\mid&\eunchecked{\overline{x}}{e}

@@

@@

There are two notions of type in \lang.  Types $\tau$ classify
word-sized values including integers and pointers, while types
$\omega$ classify multi-word values such as arrays, null-terminated
arrays, functions, and single-word-size values.
%
Pointer types ($\tptr{\omega}{m}$) include a mode annotation ($m$)
which is either checked (\cmode) or unchecked (\umode) and a type
($\omega$) denoting valid values that can be pointed to. Array types include both the type of
elements ($\tau$) and a bound ($\bvar$) comprised of an upper and
lower bound on the size of the array ($(b_l,b_h)$). Bounds $b$ are
limited to integer literals $n$ and expressions $x + n$.
Whether an array pointer is null terminated or not is determined by annotation
$\kappa$, which is $nt$ for null-terminated arrays, and $\cdot$
otherwise (we elide the $\cdot$ when writing the type).
\systemname function types ($\tfun{\overline{x}}{\overline{\tau}}{\tau}$)
reflect its dependent function declrations,
where $\overline{x}$ represents
a list of \tint type variables in a dependent function header,
which bind all bounds appearing in $\overline{\tau}$ and $\tau$.
We have a well-formed requirement for a function type,
such that all variables in $\overline{\tau}$ and $\tau$ are bounded by $\overline{x}$.
Here is the
corresponding \checkedc syntax for these types:
\[\hspace*{-0.5em}
\begin{array}{l}
\begin{array}{rcl}
$\code{array_tptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tarrayptr{0}{n}{\tau}{\tmode}
\\[0.2em]
$\code{nt_array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tntarrayptr{0}{n}{\tau}{\cmode}
\end{array}
\\[0.2em]
$\code{tptr<(int)(nt_array_tptr<}$\tau$\code{> : count(}$n$\code{),}$\\
\qquad\qquad$\code{nt_array_tptr<}$\tau$\code{>)>: count(}$n$\code{))>}$
\\[0.2em]
\Leftrightarrow\;\; $\tptr{(\tfun{n}{ \tntarrayptr{0}{n}{\tau}{\tmode} * \tntarrayptr{0}{n}{\tau}{\tmode}}{\tint})}{\tmode}$
\end{array}
\]
As a convention we write $\tptr{\tarrayb{b}{\tau}}{\cmode}$ to mean
$\tptr{\tarray{0}{b}{\tau}}{\cmode}$, so the above examples could
be rewritten $\tptr{\tarrayb{n}{\tau}}{\cmode}$ and
$\tptr{\tntarrayb{n}{\tau}}{\cmode}$, respectively.

\lang expressions include literals ($n\!:\!\tau$), variables ($x$), memory
allocation ($\emalloc{\omega}$), let binding ($\elet{x}{e_1}{e_2}$),
static casts ($\ecast{\tau}{e}$), dynamic casts ($\edyncast{\tau}{e}$)
(assumed at compile-time and verified at run-time, see Sec.~\ref{sec:example}),
function calls ($\ecall{f}{\overline{e}}$), addition
($\ebinop{e_1}{e_2}$), pointer dereference and assignment ($\estar{e}$)
and ($\eassign{e_1}{e_2}$), resp.), unchecked blocks ($\eunchecked{e}$),
the \texttt{strlen} operation ($\estrlen{x}$), and conditionals ($\eif{e}{e_1}{e_2}$).
% \review{III.A.: the "dynamic cast" terminology may be briefly confused for C++'s
%   RTTI-based dynamic cast feature}
% \mwh{Added reference back to Section 2-B}

@@

\lang aims to be simple enough to work with, but powerful enough to
encode realistic \checkedc idioms. For example, mutable local
variables can be encoded as immutable locals that point to the heap;
the use of \code{&} can be simulated with \code{malloc};
and loops can be encoded as recursive function calls. \code{struct}s are
not in Fig.~\ref{fig:checkc-syn} for space reasons, but they are
actually in our model, and developed in
\iftr
Appendix~\ref{appx:struct}.
\else
the supplemental report~\cite{checkedc-tech-report}.
\fi
C-style \code{union}s have no safe typing
in \checkedc, so we omit them. By default, functions are assumed to
be within checked regions; placing the body in an \code{unchecked}
expression relaxes this, and within that, checked regions can be
nested via function calls. Bounds are restricted slightly:
rather than allowing arbitrary sub-expressions, bounds must be either
integer literals or variables plus an integer offset, which accounts
for most uses of \code{bounds} in \checkedc programs. \lang bounds are
defined as relative offsets, not absolute ones, as in the second part
of Fig.~\ref{fig:checkedc-example}. We see no technical problem to
modeling absolute bounds, but it would be a pervasive change so we
have not done so.

  We have mechanized two models of \lang, one in Coq
  and one in PLT Redex~\cite{pltredex}, which is a semantic
  engineering framework implemented in Racket. Redex provides direct
  support for specifying the operational semantics and typing with logical
  rules, but then automatically makes them executable and subject to
  randomized testing, which is very useful during development.
  The model we present in the paper faithfully
  represents both mechanizations, but there are some differences for
  presentation purposes. For example, the paper and the Coq model use an explicit
  stack, whereas the Redex model uses \code{let} bindings to simulate
  one (simplifying term generation for randomized
  testing).
\iftr
  Appendix~\ref{app:model-diffs} 
\else
  The supplemental report~\cite{checkedc-tech-report}
\fi
  outlines the differences
  between the two models and the paper formalism.

% \mwh{NOTE: COULD WORK THE FOLLOWING INTO THE ABOVE DESCRIPTION OF
%   POINTERS: Array and NT-array types have two relative bounds, whose structures
% can be either an integer or a variable plus an integer. For example,
% if we have the expression
% $x\texttt{=}\emalloc{\tarrayb{(0,10)}{\tint}}$, $x$ then has the type
% $\tptr{\tarrayb{10}{\tint}}{\cmode}$, which represents an array
% pointer of size $10$. The bounds ($(b_l,b_h)$ in
% $\evalue{p}{\tptr{\tarrayb{(b_l,b_h)}{\tau}}{m}}$) indicate relative
% offsets from this pointer of the accessible memory, i.e., $p+b_l$ and
% $p+b_h$. As such, a pointer can only be directly dereferenced if 0 is
% included within the annotated range.  If following the $\emalloctext$
% operation, we execute $*(x-1)$ and $*(x\plus 10)$. The two expressions
% are not valid, because the type of the two expressions $(x-1)$ and
% $(x\plus 10)$ are $\tarrayptr{1}{11}{\tint}{\cmode}$ and
% $\tarrayptr{-10}{0}{\tint}{\cmode}$ and $0$ in these two cases are not
% in the ranges: $[1,11)$ and $[-10,0)$. Thus, in order for a $\cmode$
% pointer with type $\tarrayptr{b_l}{b_h}{\tint}{\cmode}$ to be
% accessible in \checkedc, $0$ must be in the range of $[b_l,b_h)$.}

\begin{figure}
{\small
$    \begin{array}{l}
    \begin{array}{lll}
\mu & ::= & \evalue{n}{\tau} \mid \bot\\
e & ::= & \ldots \mid \ret{x}{\mu}{e}\\
r & ::= & e \mid \enull \mid \ebounds\\
E &::=& \Box \mid \elet{x}{E}{e} \mid \ecall{f}{\overline{E}}\mid\ecast{\tau}{E} \mid \edyncast{\tau}{E} \\[0.2em]
&& \mid \ebinop{\evalue{n}{\tau}}{E}\mid\estar{E}\mid\eassign{E}{e} \\[0.2em]
&&\mid\eassign{\evalue{n}{\tau}}{E}\mid\eunchecked{E}\mid\eif{E}{e}{e}\\[0.2em]
\overline{E} &::=& E \mid  \evalue{n}{\tau},\overline{E} \mid \overline{E}, e
\end{array}
\\ \\
    \end{array} 
$
  \begin{mathpar}
    \inferrule{ m=\mode(E) \\
      e=E[e'] \\
      (\varphi,\heap,e') \longrightarrow (\varphi',\heap',e'')}
    {(\varphi,\heap,e)\longrightarrow_{m} (\varphi',\heap',E[e''])}

@@

  \end{mathpar}
}
  \caption{\lang Semantics: Evaluation}
  \label{fig:c-context}
\end{figure}

@@

The operational semantics for \lang is defined as a small-step
transition relation with the judgment $ (\varphi,\heap,e)
\longrightarrow_m (\varphi',\heap',r)$. Here, $\varphi$ is a
\emph{stack} mapping from variables to values $\evalue{n}{\tau}$ and
$\heap$ is a \emph{heap} mapping addresses (integer literals) to
values $\evalue{n}{\tau}$; for both we ensure
$\fv(\tau)=\emptyset$.
While heap bindings can change, stack bindings are immutable---once
variable $x$ is bound to $\evalue{n}{\tau}$ in $\varphi$, that binding will not
be updated; we can model mutable stack variables as pointers into the
mutable heap.
As mentioned, value $\evalue{0}{\tau}$
represents a $\enull$ pointer when $\tau$ is a pointer type;
correspondingly, $\heap(0)$ should always be undefined.
The relation steps to a \emph{result} $r$,
which is either an expression or a $\enull$ or $\ebounds$ failure,
representing a null-pointer dereference or out-of-bounds access,
respectively. Such failures are a \emph{good} outcome; stuck states
(non-value expressions that cannot transition to a result $r$)
characterize undefined behavior.
%
The mode $m$ indicates whether the
stepped redex within $e$ was in a checked ($\cmode$) or
unchecked ($\umode$) region.

The rules for the main operational semantics
judgment---\emph{evaluation}---are given at the bottom of
Fig.~\ref{fig:c-context}.
The first rule takes an expression $e$, decomposes
it into an \emph{evaluation context} $E$ and a sub-expression $e'$
(such that replacing the hole $\Box$ in $E$ with $e'$ would yield
$e$), and then evaluates $e'$ according to the \emph{computation}
  relation $(\varphi,\heap,e') \longrightarrow (\varphi,\heap,e'')$,
whose rules are given in Fig.~\ref{fig:semantics}, discussed
shortly. 
The second rule handles conditionals $\eif{\estar{x}}{e_2}{e_3}$ in
redex position specially, delegating directly to the \textsc{S-IfNTT} computation
rule, which supports bounds widening; we discuss this rule
shortly. When the second and first rules could both apply, we always
prefer the second.\footnote{This approach is that of the PLT Redex model of \lang; the Coq
development uses a slightly simpler syntax to achieve the same
effect.}
% \review{the special case raises questions, e.g. why is this syntax-driven and
%   not type-driven? }
% \liyi{This describes the semantic transition rules. We are using context evaluation framework to define the transition rules as the $E$ definition in Fig.3. like $\frac{x \Rightarrow y}{x+z \Rightarrow y + z}$, I don't know how type-driven can help us define translation rules.  }
% \mwh{Don't follow the above. I don't see this ``context transition
%   rule'' anywhere, and I'm not sure how it would fire, if we had it.}
% \liyi{The comment seems to confuse the meaning of the text about the if-then-else rules. Making the rule specific will help. }
The $\mode$ function
determines the mode when evaluating $e'$ based on the context $E$:
if the $\Box$ occurs within $(\eunchecked{E'})$ inside $E$, then the mode is
$\umode$; otherwise, it is $\cmode$. Evaluation contexts $E$ define a
standard left-to-right evaluation order. (We explain the
$\ret{x}{\mu}{e}$ syntax shortly.)
% For a term $e$, $C$ is a context if any
% only if it contains a $\Box$ hole term in one of the subterm position
% in $e$. $\Box$ indicates the break point where we split $e$ into $C$
% and a subterm $e'$ such that $C[e'] = e$.  For a function call
% $\ecall{f}{\overline{e}}$, a valid context ($\overline{C}$) is defined
% as any one of the elements in the list $\overline{e}$ being a
% context. We show a rule of the upper level \checkedc semantics. The
% rule splits a term $e$ into a context $C$ and redex $e'$.  If subterm
% $e'$ is transitioned to $e_a$, $C[e']$ is transitioned to
% $C[e_a]$.

\begin{DIFnomarkup}
\begin{figure*}[t]
{\small
  \begin{mathpar}
        \inferrule[S-Def]{\heap(n)=\evalue{n_a}{\tau_a} }
    {(\varphi,\heap,\estar{\evalue{n}{\tptr{\tau}{m}}}) \longrightarrow (\varphi,\heap,\evalue{n_a}{\tau})}

@@

        \inferrule[S-DefNTArray]{\heap(n)=\evalue{n_a}{\tau_a} \\ 0 \in [n_l,n_h]}
    {(\varphi,\heap,\estar{\evalue{n}{\tntarrayptr{n_l}{n_h}{\tau}{\cmode}}}) \longrightarrow (\varphi,\heap,\evalue{n_a}{\tau})}

>>>>

\begin{figure}
  \small \centering
  \[ \hspace*{-1.2em}
\begin{array}{l}
\begin{array}{ll}
       \text{Variables:}~ x
& \text{Integers:}~n::=\mathbb{Z} 
\end{array}
\\[0.5em]

@@

\text{Expression:}& e & ::= & 
\evalue{n}{\tau} \mid x \mid \ebinop{e}{e}\mid \ecast{\tau}{e} \mid \edyncast{\tau}{e}  \\[0.2em]
&&\mid& \estrlen{x} \mid\estar{e}\mid\eassign{e}{e}  \\[0.2em]
&&\mid& \elet{x}{e}{e} \mid \eif{e}{e}{e}
\\[0.2em]
&&\mid&  \emalloc{\xi}{\omega} \mid \ecall{e}{\overline{e}}
\\[0.2em]
&&\mid&\eunchecked{\overline{x}}{e}
\mid \echecked{\overline{x}}{e}

@@

\begin{figure}[t]
{\small
  \begin{mathpar}

  \inferrule[]
  {}
  {m \vdash \tint}

  \inferrule[]
  {\xi \wedge m\vdash \tau \\ \xi \le m}
  {m \vdash \tptr{\tallarrayb{\bvar}{\tau}}{\xi}}

  \inferrule[]
  {\xi \wedge m \vdash \tau\\ \xi \le m}
  {m \vdash \tptr{\tau}{\xi}}

  \inferrule[]
  {\xi \wedge m \vdash \tau\\ \xi \le m \\\\ \fv(\overline{\tau})\cup\fv(\tau)\subseteq \overline{x}}
  {m \vdash \tptr{(\tfun{\overline{x}}{\overline{\tau}}{\tau}}{\xi})}
  \end{mathpar}
}
{\footnotesize
\[
\begin{array}{l} 
\tmode \wedge \cmode = \umode \qquad \xi \wedge \umode = \umode
\qquad \cmode \wedge m = m 
\qquad  m_1 \wedge m_2 = m_2 \wedge m_1
\\[0.2em]
\xi \le \xi \qquad \tmode \le \xi
\end{array}
\\
\]
}
 \caption{Well-formedness for Types}
\label{fig:wftypes}
\end{figure}

@@

There are two notions of type in \lang.  Types $\tau$ classify
word-sized values including integers and pointers, while types
$\omega$ classify multi-word values such as arrays, null-terminated
arrays, functions, and single-word-size values.
%
Pointer types ($\tptr{\omega}{m}$) include a mode annotation ($m$)
which is either checked (\cmode) or unchecked (\umode) and a type
($\omega$) denoting valid values that can be pointed to. Array types include both the type of
elements ($\tau$) and a bound ($\bvar$) comprised of an upper and
lower bound on the size of the array ($(b_l,b_h)$). Bounds $b$ are
limited to integer literals $n$ and expressions $x + n$.
Whether an array pointer is null terminated or not is determined by annotation
$\kappa$, which is $nt$ for null-terminated arrays, and $\cdot$
otherwise (we elide the $\cdot$ when writing the type).
\systemname function types ($\tfun{\overline{x}}{\overline{\tau}}{\tau}$)
reflect its dependent function declrations,
where $\overline{x}$ represents
a list of \tint type variables in a dependent function header,
which bind all bounds appearing in $\overline{\tau}$ and $\tau$.
We have a well-formed requirement for a function type,
such that all variables in $\overline{\tau}$ and $\tau$ are bounded by $\overline{x}$.
Here is the
corresponding \checkedc syntax for these types:
\[\hspace*{-0.5em}
\begin{array}{l}
\begin{array}{rcl}
$\code{array_tptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tarrayptr{0}{n}{\tau}{\tmode}
\\[0.2em]
$\code{nt_array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tntarrayptr{0}{n}{\tau}{\cmode}
\end{array}
\\[0.2em]
$\code{tptr<(int)(nt_array_tptr<}$\tau$\code{> : count(}$n$\code{),}$\\
\qquad\qquad$\code{nt_array_tptr<}$\tau$\code{>)>: count(}$n$\code{))>}$
\\[0.2em]
\Leftrightarrow\;\; $\tptr{(\tfun{n}{ \tntarrayptr{0}{n}{\tau}{\tmode} \times \tntarrayptr{0}{n}{\tau}{\tmode}}{\tint})}{\tmode}$
\end{array}
\]
As a convention we write $\tptr{\tarrayb{b}{\tau}}{\cmode}$ to mean
$\tptr{\tarray{0}{b}{\tau}}{\cmode}$, so the above examples could
be rewritten $\tptr{\tarrayb{n}{\tau}}{\cmode}$ and
$\tptr{\tntarrayb{n}{\tau}}{\cmode}$, respectively.

\lang expressions include literals ($n\!:\!\tau$), variables ($x$),
 addition ($\ebinop{e_1}{e_2}$), static casts ($\ecast{\tau}{e}$), 
dynamic casts ($\edyncast{\tau}{e}$) \footnote{assumed at compile-time and verified at run-time, see \Cref{app:main}},
the \texttt{strlen} operation ($\estrlen{x}$),
pointer dereference and assignment ($\estar{e}$)
and ($\eassign{e_1}{e_2}$), resp.),
let binding ($\elet{x}{e_1}{e_2}$),
conditionals ($\eif{e}{e_1}{e_2}$),
memory allocation ($\emalloc{\xi}{\omega}$), 
function calls ($\ecall{e}{\overline{e}}$),
unchecked blocks ($\eunchecked{\overline{x}}{e}$), and checked blocks ($\echecked{\overline{x}}{e}$).
% \review{III.A.: the "dynamic cast" terminology may be briefly confused for C++'s
%   RTTI-based dynamic cast feature}
% \mwh{Added reference back to Section 2-B}

@@

Compared to the former \checkedc model \cite{li22checkedc},
there are four differences.
First, the \systemname type annotations have well-formed restrictions in \Cref{fig:wftypes}, 
for maintaining non-exposure.
Mainly, in a nested pointer $\tptr{(... \tptr{\tau}{\xi'} ...)}{\xi}$, if the outside mode $\xi$ is 
$\tmode$ or $\umode$, the inside mode $\xi'$ cannot be $\cmode$.
It is worth noting that pointer modes are a three point partial order ($\le$),
where $\tmode$ is the infimum, and
$\xi\wedge m$ is a special meet operation that projects pointer modes onto context modes,
such that a pointer mode $\tmode$ is projected as $\umode$.
Second, $\emalloc{\xi}{\omega}$ includes a mode flag
$\xi$ for allocating different pointers in different heap mode regions.
For simplicity, we disallow $\omega$ to be a function type ($\tfun{\overline{x}}{\overline{\tau}}{\tau}$),
Third, the first expression $e$ in a function call ($\ecall{e}{\overline{e}}$) represents a function pointer.
Fourth, we update $\eunchecked{\overline{x}}{e}$ 
to specify $\overline{x}$ restricting all free variables appearing in $e$.
Additionally, we create $\echeckedtext$ blocks indicating the context mode change from $\umode$ to $\cmode$.
Both the free variables in the $\euncheckedtext$ and $\echeckedtext$ blocks
and the return types cannot be $\cmode$ mode for maintaining non-exposure,
i.e., no checked pointer can be communicated between $\euncheckedtext$ and $\echeckedtext$ blocks.

\lang aims to be simple enough to work with, but powerful enough to
encode realistic \checkedc idioms. For example, mutable local
variables can be encoded as immutable locals that point to the heap;
the use of \code{&} can be simulated with \code{malloc};
and loops can be encoded as recursive function calls. \code{struct}s are
not in Fig.~\ref{fig:checkc-syn} for space reasons, but they are
actually in our model, and developed in
\iftr
Appendix~\ref{appx:struct}.
\else
the supplemental report~\cite{checkedc-tech-report}.
\fi
C-style \code{union}s have no safe typing
in \checkedc, so we omit them.
Additional syntactic descrition is listed in the \checkedc formalism paper \cite{li22checkedc}.

\ignore{
 By default, functions are assumed to
be within checked regions; placing the body in an \code{unchecked}
expression relaxes this, and within that, checked regions can be
nested via function calls. Bounds are restricted slightly:
rather than allowing arbitrary sub-expressions, bounds must be either
integer literals or variables plus an integer offset, which accounts
for most uses of \code{bounds} in \checkedc programs. \lang bounds are
defined as relative offsets, not absolute ones, as in the second part
of Fig.~\ref{fig:checkedc-example}. We see no technical problem to
modeling absolute bounds, but it would be a pervasive change so we
have not done so.

  We have mechanized two models of \lang, one in Coq
  and one in PLT Redex~\cite{pltredex}, which is a semantic
  engineering framework implemented in Racket. Redex provides direct
  support for specifying the operational semantics and typing with logical
  rules, but then automatically makes them executable and subject to
  randomized testing, which is very useful during development.
  The model we present in the paper faithfully
  represents both mechanizations, but there are some differences for
  presentation purposes. For example, the paper and the Coq model use an explicit
  stack, whereas the Redex model uses \code{let} bindings to simulate
  one (simplifying term generation for randomized
  testing).
\iftr
  Appendix~\ref{app:model-diffs} 
\else
  The supplemental report~\cite{checkedc-tech-report}
\fi
  outlines the differences
  between the two models and the paper formalism.
}
% \mwh{NOTE: COULD WORK THE FOLLOWING INTO THE ABOVE DESCRIPTION OF
%   POINTERS: Array and NT-array types have two relative bounds, whose structures
% can be either an integer or a variable plus an integer. For example,
% if we have the expression
% $x\texttt{=}\emalloc{\tarrayb{(0,10)}{\tint}}$, $x$ then has the type
% $\tptr{\tarrayb{10}{\tint}}{\cmode}$, which represents an array
% pointer of size $10$. The bounds ($(b_l,b_h)$ in
% $\evalue{p}{\tptr{\tarrayb{(b_l,b_h)}{\tau}}{m}}$) indicate relative
% offsets from this pointer of the accessible memory, i.e., $p+b_l$ and
% $p+b_h$. As such, a pointer can only be directly dereferenced if 0 is
% included within the annotated range.  If following the $\emalloctext$
% operation, we execute $*(x-1)$ and $*(x\plus 10)$. The two expressions
% are not valid, because the type of the two expressions $(x-1)$ and
% $(x\plus 10)$ are $\tarrayptr{1}{11}{\tint}{\cmode}$ and
% $\tarrayptr{-10}{0}{\tint}{\cmode}$ and $0$ in these two cases are not
% in the ranges: $[1,11)$ and $[-10,0)$. Thus, in order for a $\cmode$
% pointer with type $\tarrayptr{b_l}{b_h}{\tint}{\cmode}$ to be
% accessible in \checkedc, $0$ must be in the range of $[b_l,b_h)$.}

\begin{figure}
{\small
$\hspace*{-1.2em}
    \begin{array}{l}
    \begin{array}{lll}
\mu & ::= & \evalue{n}{\tau}\\
e & ::= & \ldots \mid \ret{x}{\mu}{e}\\
r & ::= & e \mid \enull \mid \ebounds\\
E &::=& \Box \mid \ebinop{E}{e} \mid \ebinop{\evalue{n}{\tau}}{E}\mid \ecast{\tau}{E} \mid \edyncast{\tau}{E} \mid\estar{E}\mid\eassign{E}{e}\\[0.2em]
&&\mid\eassign{\evalue{n}{\tau}}{E}\mid \elet{x}{E}{e}\mid \eif{E}{e}{e}\\[0.2em]
&&\mid \ecall{E}{\overline{e}} \mid \ecall{\evalue{n}{\tau}}{\overline{E}} \mid 
\eunchecked{\overline{x}}{E}
\mid \echecked{\overline{x}}{E}

\end{array}
\\ \\
    \end{array} 
$
  \begin{mathpar}
    \inferrule{ m=\mode(E) \\
      e=E[e'] \\
      (\varphi,\heap,e') \longrightarrow (\varphi',\heap',e'')}
    {(\varphi,\heap,e)\longrightarrow_{m} (\varphi',\heap',E[e''])}

@@

    \inferrule{ \umode=\mode(E) \\
      e=E[e'] \\
      \tau=\type(e')}
    {(\varphi,\heap,e)\longrightarrow_{\umode} (\varphi,\heap,E[\evalue{0}{\tau}])}

  \end{mathpar}
}
{\footnotesize
\[
\begin{array}{l} 
\mode(E)=\mode'(E,\cmode)
\\[0.2em]
\mode'(\Box,m)=m
\\[0.2em]
\mode'(\eunchecked{\overline{x}}{E},m)=\mode'(E,\umode)
\\[0.2em]
\mode'(\echecked{\overline{x}}{E},m)=\mode'(E,\cmode)
\\[0.2em]
\mode'(\alpha({E}),m)=\mode'(E,m)\;\;{[\emph{owise}]}
\end{array}
\\
\]
}
  \caption{\lang Semantics: Evaluation}
  \label{fig:c-context}
\end{figure}

@@

The operational semantics for \lang is defined as a small-step
transition relation with the judgment $ (\varphi,\heap,e)
\longrightarrow_m (\varphi',\heap',r)$. Here, $\varphi$ is a
\emph{stack} mapping from variables to values $\evalue{n}{\tau}$ and
$\heap$ is a \emph{heap} that is partitioned into two parts ($\cmode$ and $\umode$ regions), each of which
maps addresses (integer literals) to values $\evalue{n}{\tau}$.
If a pointer has mode $\cmode$, it lives in the $\cmode$ region; otherwise, it lives in the $\umode$ region.
We wrote $\heap(m,n)$ to retrieve the $n$-location heap value in the $m$ region,
and $\heap(m)[n\mapsto \mu]$ to update location $n$ with the value $\mu$.
It is worth noting that \systemname is not a fat-pointer system;
thus, in every heap update, the value type annotation remains the same through program executions.
Additionally, for both stack and heap, 
we ensure $\fv(\tau)=\emptyset$ for all the value type annotations $\tau$.

While heap bindings can change, stack bindings are immutable---once
variable $x$ is bound to $\evalue{n}{\tau}$ in $\varphi$, that binding will not
be updated; we can model mutable stack variables as pointers into the
mutable heap.
As mentioned, value $\evalue{0}{\tau}$
represents a $\enull$ pointer when $\tau$ is a pointer type;
correspondingly, $\heap(0)$ should always be undefined.
The relation steps to a \emph{result} $r$,
which is either an expression or a $\enull$ or $\ebounds$ failure,
representing a null-pointer dereference or out-of-bounds access,
respectively. Such failures are a \emph{good} outcome; stuck states
(non-value expressions that cannot transition to a result $r$)
characterize undefined behavior.
%
The context mode $m$ indicates whether the
stepped redex within $e$ was in a checked ($\cmode$) or
unchecked ($\umode$) region.

The rules for the main operational semantics
judgment---\emph{evaluation}---are given at the bottom of
Fig.~\ref{fig:c-context}.
The first rule takes an expression $e$, decomposes
it into an \emph{evaluation context} $E$ and a sub-expression $e'$
(such that replacing the hole $\Box$ in $E$ with $e'$ would yield
$e$), and then evaluates $e'$ according to the \emph{computation}
  relation $(\varphi,\heap,e') \longrightarrow (\varphi,\heap,e'')$,
whose rules are given in Fig.~\ref{fig:semantics}, discussed
shortly. 
The second rule handles conditionals $\eif{\estar{x}}{e_2}{e_3}$ in
redex position specially, delegating directly to the \textsc{S-IfNTT} computation
rule, which supports bounds widening; we discuss this rule
shortly. When the second and first rules could both apply, we always prefer the second.
The third rule describes the exception handling 
for possible crashing behaviors in unchecked region.
A program in $\umode$ mode can non-deterministically crash
and the \systemname sandbox mechanism recorvers
the program to a safe point ($\evalue{0}{\tau}$)
and continues with the existing program state.

\begin{DIFnomarkup}
 \begin{figure}[t]
 {\small

 \begin{mathpar}
   \inferrule
       {}
       {m;\Theta;\heap;\sigma \vdash n : \tint}

   \inferrule
       {}
       {m;\Theta;\heap;\sigma \vdash n : \tptr{\omega}{\umode}}

   \inferrule
       {}
       {m;\Theta;\heap;\sigma \vdash 0 : \tptr{\omega}{\xi}}

   \inferrule
       {(\evalue{n}{\tptr{\omega}{\xi}})\in \sigma}
       {m;\Theta;\heap;\sigma \vdash n : \tptr{\omega}{\xi}}

   \inferrule
       {\tptr{\omega'}{\xi'} \sqsubseteq_{\Theta} \tptr{\omega}{\xi} 
            \\ m;\Theta;\heap;\sigma \vdash n : \tptr{\omega'}{\xi'}}
       {m;\Theta;\heap;\sigma \vdash n : \tptr{\omega}{\xi}}

   \inferrule
       { \xi \le m 
     \\\Xi(m,n)=\tau\;(\evalue{\overline{x'}}{\overline{\tau}})\;(\xi,e)
       \\  \overline{x} = \{x|(x:\tint) \in (\overline{x'}:\overline{\tau}) \}}
       {m;\Theta;\heap;\sigma \vdash n : \tptr{(\tfun{\overline{x}}{\overline{\tau}}{\tau})}{\xi}}

   \inferrule
       {\neg\funptr(\omega)\\ \xi \le m\\
        \forall i \in [0,\size(\omega)) \;.\;
            m;\Theta;\heap;(\sigma \cup \{(n:\tptr{\omega}{\xi})) \}\vdash \heap(m,n+i)}
       {m;\Theta;\heap;\sigma \vdash n : \tptr{\omega}{\xi}}
 \end{mathpar}
 }
{\footnotesize
\[
\begin{array}{l} 
\funptr(\tfun{\overline{x}}{\overline{\tau}}{\tau}) = \texttt{true}
\qquad
\funptr(\omega) = \texttt{false}\;\;{[\emph{owise}]}
\end{array}
\\
\]
}
 \caption{Verification/Type Rules for Constants}
 \label{fig:const-type}
 \end{figure}
\end{DIFnomarkup}

%\footnote{This approach is that of the PLT Redex model of \lang; the Coq
%development uses a slightly simpler syntax to achieve the same
%effect.}
% \review{the special case raises questions, e.g. why is this syntax-driven and
%   not type-driven? }
% \liyi{This describes the semantic transition rules. We are using context evaluation framework to define the transition rules as the $E$ definition in Fig.3. like $\frac{x \Rightarrow y}{x+z \Rightarrow y + z}$, I don't know how type-driven can help us define translation rules.  }
% \mwh{Don't follow the above. I don't see this ``context transition
%   rule'' anywhere, and I'm not sure how it would fire, if we had it.}
% \liyi{The comment seems to confuse the meaning of the text about the if-then-else rules. Making the rule specific will help. }
The $\mode$ function
determines the context mode when evaluating $e'$ based on the context $E$.
For any program execution, the top context mode always starts with $\cmode$ ($\mode(E)=\mode'(E,\cmode)$).
The result context mode is depeneded on where the $\Box$ locates.
If it occurs within $(\eunchecked{\overline{x}}{E'})$ inside $E$ without
a surrounding $\echeckedtext$ block, then the mode is
$\umode$; if the exact opposite happens (within $(\echecked{\overline{x}}{E'})$ without surrounding $\euncheckedtext$ blocks),
the mode is $\cmode$; the other cases ($\mode'(\alpha({E}),m)=\mode'(E,m)$) are recursively defined by the above base cases ($\alpha$ represents any syntactic categories in \Cref{fig:c-context} other than $\Box$, $\euncheckedtext$, and $\echeckedtext$).
Evaluation contexts $E$ define a standard left-to-right evaluation order. (We explain the
$\ret{x}{\mu}{e}$ syntax shortly.)
% For a term $e$, $C$ is a context if any
% only if it contains a $\Box$ hole term in one of the subterm position
% in $e$. $\Box$ indicates the break point where we split $e$ into $C$
% and a subterm $e'$ such that $C[e'] = e$.  For a function call
% $\ecall{f}{\overline{e}}$, a valid context ($\overline{C}$) is defined
% as any one of the elements in the list $\overline{e}$ being a
% context. We show a rule of the upper level \checkedc semantics. The
% rule splits a term $e$ into a context $C$ and redex $e'$.  If subterm
% $e'$ is transitioned to $e_a$, $C[e']$ is transitioned to
% $C[e_a]$.

\begin{DIFnomarkup}
\begin{figure*}[t]
{\small
  \begin{mathpar}
        \inferrule[S-DefC]{\heap(\cmode,n)=\evalue{n_a}{\tau_a} }
    {(\varphi,\heap,\estar{\evalue{n}{\tptr{\tau}{c}}}) \longrightarrow (\varphi,\heap,\evalue{n_a}{\tau})}

    \inferrule[S-AssignArrC]{\heap(\cmode,n)=\evalue{n_a}{\tau_a}\\ 0 \in [n_l,n_h) }
      {(\varphi,\heap,\eassign{\evalue{n}{\tallarrayptr{n_l}{n_h}{\tau}{\cmode}}}{\evalue{n_1}{\tau_1}}) \longrightarrow (\varphi,(\heap,\cmode)[n \mapsto \evalue{n_1}{\tau_a}],\evalue{n_1}{\tau})}

        \inferrule[S-DefT]{\heap(\umode,n)=\evalue{n_a}{\tau_a}
         \\  \umode;\emptyset;\heap ; \emptyset \vdash\evalue{n_a}{\tau} }
    {(\varphi,\heap,\estar{\evalue{n}{\tptr{\tau}{\tmode}}}) \longrightarrow (\varphi,\heap,\evalue{n_a}{\tau})}

    \inferrule[S-AssignArrT]{\heap(\umode,n)=\evalue{n_a}{\tau_a}\\ 0 \in [n_l,n_h)
               \\  \umode;\emptyset;\heap ; \emptyset \vdash\evalue{n_1}{\tau} }
      {(\varphi,\heap,\eassign{\evalue{n}{\tallarrayptr{n_l}{n_h}{\tau}{\tmode}}}{\evalue{n_1}{\tau_1}}) \longrightarrow (\varphi,(\heap,\umode)[n \mapsto \evalue{n_1}{\tau_a}],\evalue{n_1}{\tau})}

@@

     \inferrule[S-IfNTTC]{\varphi(x)=\evalue{n}{\tntarrayptr{n_l}{0}{\tau}{\cmode}} \\ \heap(\cmode,n)\neq 0}
    {(\varphi,\heap,\eif{*x}{e_1}{e_2}) \longrightarrow (\varphi[x\mapsto \evalue{n}{\tntarrayptr{n_l}{1}{\tau}{\cmode}}],\heap,e_1)}
